Tempus: Probabilistic Network Latency Verifier

### 2022-10-06 paper/main.tex
[Paper] expanded the verification part

\begin{document}
\begin{abstract}
    To combat problems and bugs that a given network might have, network verifiers have emerged as 
    one of the promising solutions. 
    State-of-the-art network verifiers mainly focused on evaluating qualitative properties under 
    various scenarios of network failures, such as reachability under k-link failure. 
    However, as modern networks evolved and performance need becomes more stringent -- often 
    expressed in terms of Service Level Agreement (SLA) -- there is a need to evolve network 
    verifiers to also reason about quantitative performance properties. 
    Works in quantitative network verifiers that has arose in recent years mainly focuses on one 
    side of the network performance metric: bandwidth and load violation properties. 
    Questions about the other side of network performance metric, latency, were left unanswered.

    In this work, we introduce a verifier framework, Tempus, that can answer the probability of 
    a given temporal properties being true given latency distributions of individual links and 
    nodes in the network. 
    Early evaluation shows that Tempus can verify bounded reachability property -- the probability
    that the average delay of packets traversing from a source to destination node is below $T$
    time unit -- by only adding a fraction of the overhead introduced by the qualitative verifier 
    its built upon.
\end{abstract}

@@

\section{Introduction}
Modern networks consisted of various distributed protocols such as OSPF and BGP that exchange 
routing information so that a packet can reach their intended destination efficiently, even in the 
event of a failure.
Due to their configuration intricacies however, these protocols are notoriously hard to get right. 
It is hard for a network engineer to reason about whether their configured network fulfilled some 
intended property in various possible states of the network. 
Thus, they are left between the choice of accepting the reality of Murphy's Law or getting a tool 
to help them in this task, namely, network control plane verifier.

@@

\section{Proposed Approach}
We will start with the assertion that latency is a property that only make sense 
after connectivity between two nodes has been established. 
In other words, if two nodes in a network aren't connected (due to physical failure 
or ACL policy), then the latency between them will \textit{always} be infinite.
Because of this, we divide the problem of latency verification into two parts: 
verifying that two nodes are reachable (\textit{functional} property) and only if 
the functional property is fulfilled, we would verify whether the latency between 
two nodes fulfilled some additional condition (\textit{temporal} property)

@@

\subsection{Latency Graph}
There are a lot of components in a network that may introduce some latency into 
a transmitting packet. 
The principal challenge for us, then, is to determine the appropriate level of abstraction; 
detailed enough in order to accomplish our verification goal in a correct and accurate manner but 
abstract enough for the problem to remain tractable. 
We settle on modeling two sources of latency that we deem significant: propagation delay and 
queuing delay.

@@

\begin{itemize}
    \item Each $v_i$ is expanded into $v_i^{in}, v_i^{out-1}, ..., v_i^{out-n} \in V_l$ 
    where $n$ is the amount of outward neighbors that $v_i$ has in $G_t$.

@@

Semantically, $v_i^{out-j}$ represents a port, normal links that are copied from $G_t$ represents 
a link with a propagation delay, and newly introduced links $e_i^j$ represents an output queue 
in node $v_i$ with a queuing delay.

\subsection{Latency Verification}
From these two graphs, we then do the verification in the following way:

\textbf{First}, we do the functional verification with $G_t$ similar to what NetDice 
\cite{steffen2020probabilistic} have done. 
In addition to that, we also collect additional information from each state (e.g. convergent 
paths) to be used in the temporal verification step. 
After the functional verifier finished exploring, we then move to the next step.

\textbf{Second}, we compute the total path latency distribution from the convergent path(s) in each 
state with $G_l$.
This is done by convolving over the latency distribution of each component in the path. 
Since not all convolutions can be calculated analytically, we implemented a numerical convolution 
method via mixture distribution, DIRECT \cite{rover2017discrete}, which is able to convolve two 
distribution with a KL divergence error bound.

\textbf{Third}, we calculate the probability of a given temporal property being true in that state. 
For example, we define \textbf{bounded reachability property} (probability of whether a packet can 
traverse from a source to destination below some time unit $T$) which can be computed by 
integrating the PDF from $0$ to $T$. 
We then combine all of the \textit{temporal} and \textit{functional} in each state to get the 
probability of a given temporal property being true in this network.

@@

On the performance side, we will evaluate the verifier by measuring the amount of additional 
overhead that the temporal verifier would introduce, measured by the amount of states that 
needs to be re-explored (in the second step) and / or the amount of convolution operation is 
performed.
% We will use two main datasets: using topologies from the topology zoo \cite{knight2011internet} 
% with a realistic assumed temporal distribution and topologies that has node measurements that 
% could be used to form a output queue distribution.

>>>>

\begin{document}
\begin{abstract}
    To combat problems and bugs that a given network might have, network verifiers have emerged as 
    one of the promising solutions. 
    State-of-the-art network verifiers mainly focused on evaluating qualitative properties under 
    various scenarios of network failures, such as reachability under k-link failure. 
    However, as modern networks evolved and performance need becomes more stringent -- often 
    expressed in terms of Service Level Agreement (SLA) -- there is a need to evolve network 
    verifiers to also reason about quantitative performance properties. 
    Works in quantitative network verifiers that has arose in recent years mainly focused on one 
    side of the network performance metric: bandwidth and load violation properties. 
    Questions about the other side of network performance metric, latency, were left unanswered.

    In this work, we introduce a verifier framework, Tempus, that can answer the probability of 
    a given temporal property being true given latency distributions of individual links and 
    nodes in the network. 
    Early evaluation shows that Tempus can verify bounded reachability property -- the probability
    that the average delay of packets traversing from a source to destination node is below $T$
    time unit -- by only adding a fraction of the state exploration overhead introduced by the 
    qualitative verifier it's built upon.
\end{abstract}

@@

\section{Introduction}
Modern networks consisted of various distributed protocols such as OSPF and BGP that exchange 
routing information so that a packet can reach their intended destination efficiently, even in the 
event of a component failure.
Due to their configuration intricacies however, these protocols are notoriously hard to get right. 
It is hard for a network engineer to reason about whether their configured network fulfilled some 
intended property in various possible states of the network. 
Thus, they are left between the choice of accepting the reality of Murphy's Law or getting a tool 
to help them in this task, namely, network control plane verifier.

@@

\section{Encoding}
We will start with the assertion that latency is a property that only make sense 
after connectivity between two nodes has been established. 
In other words, if two nodes in a network aren't connected (due to physical failure 
or ACL policy), then the latency between them will \textit{always} be infinite.
Because of this, we divide the problem of latency verification into two parts: 
verifying that two nodes are reachable (\textit{functional} property) and only if 
the functional property is fulfilled, we would verify whether the latency between 
two nodes fulfilled some additional condition (\textit{temporal} property)

@@

\subsection{Latency Graph}
There are many source of latency in a network. 
The principal challenge for us, then, is to determine the appropriate level of abstraction; 
detailed enough in order to accomplish our verification goal correctly but 
abstract enough for the problem to remain tractable. 
We settle on modeling two significant sources of latency: propagation and 
queuing delay.

@@

\begin{itemize}
    \item Each $v_i$ is expanded into $v_i^{in}, v_i^{out-1}, ..., v_i^{out-n} \in V_l$ 
    where $n$ is the amount $v_i$'s outward neighbors.

@@

Semantically, $v_i^{out-j}$ represents a port, normal links that are copied from $G_t$ represents 
a link that has a propagation delay, and newly introduced links $e_i^j$ represents an output queue 
in node $v_i$ that has a queuing delay.

\section{Verification}
From these two graphs, we then do the verification in the following way:

\subsubsection{Functional Verification}
Let a network state $s_i$ be defined as a 2-tuple $(U, D)$ where $U$ is a set of links that are 
alive / up and $D$ is a set of links that are dead / down.
We then define a set $S$ that represents all possible network states in a given topology. 
Our goal is to identify a subset of $S$ where each element is a network state that can connect 
the source and destination node given a convergent behavior of a routing protocol.

This framework fits nicely with the framework of probability theory, where $S$ represents the 
sample space and the power set of $S$ represents the possible events, one of which is the one
we care about: the event that a packet from a source node can reach a destination node.
The way we assign probability to events is to simply do a summation over the product of all the 
failure and success rate of each links in the network state.

The straightforward and naive way to compute the probability of this event is to iterate over all 
the network state. 
We then determine whether that network state can fulfill said property and which path it will 
take. 
If yes, then we would compute the probability from the product of the link's failure and success 
rate, and add it to the running sum.
If no, then we could skip the computation altogether.

While natural, this brute-force algorithm is obviously doesn't scale since it will take $2^n$ 
iteration ($n$ as the number of links in the topology). 
NetDice \cite{steffen2020probabilistic} solves this issue by noticing that we could skip over some 
events that are guaranteed not to change the convergent paths of the current network state, by 
marginalizing over \textit{cold edges}.

In NetDice, we start from a perfect network state (i.e. $s_i = (U, D)$ where $D = \emptyset$).
We then determine whether that network state can fulfill said property and which path it will 
take, similar to the brute-force method.
If no, then it is safe to say that the source node can never reach the destination node in all 
network state.
If yes, then we could use the logic of the routing protocol to determine the set of cold edges, 
(i.e. edges whose failure won't change the convergent paths). 
Edges that are not part of this set is called \textit{hot edges} $h_i = (U, D)$.
Instead of using $s_i$ to compute the probability of the current state, NetDice only use the set 
of hot edges to compute the probability of them being alive ($h_i = (U, D)$ where $D = \emptyset$).
This in effect computes the probability of multiple network state at once.

Next, instead of examining other network state at random, we instead iterate over the links in 
$h_i$.
By definition, if we fail a link that belongs to $h_i$, then the convergent paths will change.
To cover all of those possibilities efficiently, NetDice iterate over all the links in $h_i$, 
disabled them, re-enabled them, and move to the next edge to do the same.
If $h_i = ((e_1, e_2, ..., e_j), \emptyset)$, then we would enqueue $(\emptyset, (e_1))$, 
$((e_1), (e_2))$, $((e_1, e_2), (e_3))$, ..., $((e_1, e_2, ..., e_{j-1}), (e_j))$.
Thus, instead of exploring $2^j$ events, we will enqueue $j$ events instead.

NetDice's exploration will effectively form a tree, where the root is the event where we have 
empty hot edges, and the leaves are where the hot edges are full.

\subsubsection{Temporal Verification}
Since we assign each links and output queue in the network with a certain latency distribution, 
computing the probability of the temporal property involves convolving the the paths we got from 
functional verification and measuring the statistical property of the resulting distribution.
However, while NetDice has shrunk the exploration space compared to brute-force iteration, for
the purpose of temporal verification, further optimization can be done.

The first is what we call \textbf{hot-edges union}. 
The idea is, while each hot-edges events has the same convergent paths, it is not guaranteed to be 
unique. 
Two or more of these events can have the same convergent paths.
Therefore, we could merge these two events and compute the convolution for a given path once.

% \textbf{First}, we do the functional verification with $G_t$ similar to what NetDice 
% \cite{steffen2020probabilistic} have done. 
% In addition, we also collect additional information from each state (e.g. convergent 
% paths) to be used in the temporal verification step.

% \textbf{Second}, we compute the total path latency distribution from the convergent path(s) in each 
% state with $G_l$.
% This is done by convolving over the latency distribution of each component in the path. 
% Since not all convolutions can be calculated analytically, we implemented a numerical convolution 
% method via mixture distribution, DIRECT \cite{rover2017discrete}, which is able to convolve two 
% distribution with a KL divergence error bound.
% Multiple states can share the same convergent path(s), so \textit{only a fraction of the states need to be 
% explored}.

% \textbf{Third}, we calculate the probability of a given temporal property being true in that state. 
% For example, given a \textit{bounded reachability property} (probability of whether a packet can 
% traverse from a source to destination below some time unit $T$), we can compute its probability by 
% integrating the PDF from $0$ to $T$. 
% We then combine all of the \textit{temporal} and \textit{functional} probability in each state to 
% get the final probability: the probability of \textit{the network} fulfilling a temporal property.

@@

On the performance side, we will evaluate the verifier by measuring the amount of additional 
overhead that the temporal verifier would introduce, measured by the amount of states that 
need to be re-explored (in the second step) and / or the amount of convolution operation performed. 
Early result on Highwinds \cite{knight2011internet} network shows that out of 2344 states 
produced by step 1, only 459 got re-explored on step 2.
